---
title: "Genetic Drift Simulation"
author: "Kevin Huang"
date: "1/8/2021"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggsci)
library(ggthemes)
```

```{r, echo=FALSE}
trajectories <- function(npops=5,popsize=20,ngen=55, p0 =0.9 ){
  sim = matrix(0,npops,ngen)
  for( pop in 1:npops) { #for each population
  sim[pop,1] = p0       #start with the allele at frequency p0 in generation 1  
  for( i in 2:ngen){    #Simulate future frequency for ngens using the binomial function
	  sim[pop,i] = rbinom(1,2*popsize, sim[pop,i-1]) / (2*popsize) 
     }
  }
  #Note the freq. each generation is probabilistically dependent on the frequency from the generation before
  ## The data are simulated; now convert to dataframe, create a column to label each population, and 
  sim <- as.data.frame(sim)
  Population <- paste("Pop", seq(1:npops), sep = " ")
  sim <- cbind(Population, sim)
  return(sim)
}
```

## Genetic Drift - Changing One Variable

The following code simulates the Wright-Fisher model. Let's see what happens with drift and fixation as we vary individual parameters.

### Initial Allele Frequency

We begin by varying the initial allele frequency. 
```{r, fig.align='center'}
#low p0
simdata <- trajectories(npops= 100, popsize = 10000, ngen = 1000, p0 = 0.1)
simdata_tidy <- pivot_longer(simdata, !Population, names_to = "Generation", names_prefix = "V", values_to = "Frequency")
simdata_tidy$Generation <- as.numeric(simdata_tidy$Generation)

low_p0 <-ggplot(simdata_tidy, aes(x = Generation, y = Frequency, group = Population)) +
  geom_line(alpha = 0.2) +xlab("Generation")
low_p0 + ggtitle("Low Initial Allele Frequency")+
  theme(plot.title = element_text(hjust = 0.5))

#mid p0
simdata <- trajectories(npops= 100, popsize = 10000, ngen = 1000, p0 = 0.5)
simdata_tidy <- pivot_longer(simdata, !Population, names_to = "Generation", names_prefix = "V", values_to = "Frequency")
simdata_tidy$Generation <- as.numeric(simdata_tidy$Generation)

mid_p0 <-ggplot(simdata_tidy, aes(x = Generation, y = Frequency, group = Population)) +
  geom_line(alpha = 0.2) +xlab("Generation")
mid_p0 + ggtitle("Mid Initial Allele Frequency")+
  theme(plot.title = element_text(hjust = 0.5))

#high p0
simdata <- trajectories(npops= 100, popsize = 10000, ngen = 1000, p0 = 0.9)
simdata_tidy <- pivot_longer(simdata, !Population, names_to = "Generation", names_prefix = "V", values_to = "Frequency")
simdata_tidy$Generation <- as.numeric(simdata_tidy$Generation)

high_p0 <-ggplot(simdata_tidy, aes(x = Generation, y = Frequency, group = Population)) +
  geom_line(alpha = 0.2) +xlab("Generation")
high_p0 + ggtitle("High Initial Allele Frequency")+
  theme(plot.title = element_text(hjust = 0.5))
```
The results are relatively predictable: At a higher initial frequency, more populations became fixed, and less became extinct. At a low initial frequency, more populations became extinct, and less became fixed. And at a middle initial frequency, both fixation and extinction were reduced. 

### Number of Generations
Next, let's vary the number of generations that our populations go through. 
```{r, fig.align='center'}
#low ngen
simdata <- trajectories(npops= 100, popsize = 10000, ngen = 100, p0 = 0.5)
simdata_tidy <- pivot_longer(simdata, !Population, names_to = "Generation", names_prefix = "V", values_to = "Frequency")
simdata_tidy$Generation <- as.numeric(simdata_tidy$Generation)

low_ngen <-ggplot(simdata_tidy, aes(x = Generation, y = Frequency, group = Population)) +
  geom_line(alpha = 0.2) +xlab("Generation")
low_ngen + ggtitle("Low Number of Generations")+
  theme(plot.title = element_text(hjust = 0.5))

#mid ngen
simdata <- trajectories(npops= 100, popsize = 10000, ngen = 1000, p0 = 0.5)
simdata_tidy <- pivot_longer(simdata, !Population, names_to = "Generation", names_prefix = "V", values_to = "Frequency")
simdata_tidy$Generation <- as.numeric(simdata_tidy$Generation)

mid_ngen <-ggplot(simdata_tidy, aes(x = Generation, y = Frequency, group = Population)) +
  geom_line(alpha = 0.2) +xlab("Generation")
mid_ngen + ggtitle("Mid Number of Generations")+
  theme(plot.title = element_text(hjust = 0.5))

#high ngen
simdata <- trajectories(npops= 100, popsize = 10000, ngen = 10000, p0 = 0.5)
simdata_tidy <- pivot_longer(simdata, !Population, names_to = "Generation", names_prefix = "V", values_to = "Frequency")
simdata_tidy$Generation <- as.numeric(simdata_tidy$Generation)

high_ngen <-ggplot(simdata_tidy, aes(x = Generation, y = Frequency, group = Population)) +
  geom_line(alpha = 0.2) +xlab("Generation")
high_ngen + ggtitle("High Number of Generations")+
  theme(plot.title = element_text(hjust = 0.5))
```
We see that when populations are given more generations, populations segregate less and go further to extinction or fixation. 

### Population Size
Finally, let's vary our population size.
```{r, fig.align='center'}
#low popsize
simdata <- trajectories(npops= 100, popsize = 100, ngen = 1000, p0 = 0.5)
simdata_tidy <- pivot_longer(simdata, !Population, names_to = "Generation", names_prefix = "V", values_to = "Frequency")
simdata_tidy$Generation <- as.numeric(simdata_tidy$Generation)

low_popsize <-ggplot(simdata_tidy, aes(x = Generation, y = Frequency, group = Population)) +
  geom_line(alpha = 0.2) +xlab("Generation")
low_popsize + ggtitle("Low Population Size")+
  theme(plot.title = element_text(hjust = 0.5))

#mid popsize
simdata <- trajectories(npops= 100, popsize = 1000, ngen = 1000, p0 = 0.5)
simdata_tidy <- pivot_longer(simdata, !Population, names_to = "Generation", names_prefix = "V", values_to = "Frequency")
simdata_tidy$Generation <- as.numeric(simdata_tidy$Generation)

mid_popsize <-ggplot(simdata_tidy, aes(x = Generation, y = Frequency, group = Population)) +
  geom_line(alpha = 0.2) +xlab("Generation")
mid_popsize + ggtitle("Mid Population Size")+
  theme(plot.title = element_text(hjust = 0.5))

#high popsize
simdata <- trajectories(npops= 100, popsize = 10000, ngen = 1000, p0 = 0.5)
simdata_tidy <- pivot_longer(simdata, !Population, names_to = "Generation", names_prefix = "V", values_to = "Frequency")
simdata_tidy$Generation <- as.numeric(simdata_tidy$Generation)

high_popsize <-ggplot(simdata_tidy, aes(x = Generation, y = Frequency, group = Population)) +
  geom_line(alpha = 0.2) +xlab("Generation")
high_popsize + ggtitle("High Population Size")+
  theme(plot.title = element_text(hjust = 0.5))
```
We can see that as we increase the population size, less and less populations go to extinction and fixation. Interestingly, at a low population size, it seems that the number of populations that either go to fixation, or go to extinction, are about equal. At a initial frequency of 50% this makes sense, as populations are initially equally as likely to go to fixation/extinction. How about at other initial frequencies?

## Genetic Drift - Changing More than One Variable 

While the previous simulations have focused on what occurs when one variable is changed, and all others remain constant, we know that this rarely happens in real populations. Let's see what happens when we change more than one variable. 

NOTE: The code is adapted from https://stephens999.github.io/fiveMinuteStats/wright_fisher_model.html

```{r, echo=FALSE, fig.align='center'}
wf_df <- data.frame()

sizes <- c(50, 100, 1000, 5000)
starting_p <- c(.01, .1, .5, .9)
n_gen <- 100
n_reps <- 100

for(N in sizes){
  for(p in starting_p){
    p0 <- p
    for(j in 1:n_gen){
      X <- rbinom(n_reps, 2*N, p)
      p <- X / (2*N)
      rows <- data.frame(replicate = 1:n_reps, N = rep(N, n_reps), 
                         Generation = rep(j, n_reps), p0 = rep(p0, n_reps), 
                         Frequency = p)
      wf_df <- bind_rows(wf_df, rows)
    }
  }
}

p <- ggplot(wf_df, aes(x = Generation, y = Frequency, group = replicate)) +
     geom_path(alpha = .5) + facet_grid(N ~ p0) + guides(colour=FALSE)
p
```
Here we see the result of varying the population size and initial frequency together. Number of generations could similarly be changed along with the other two variables; this was not included for conciseness (64 graphs would have been produced), but the code may be manipulated to show the result of changing number of generations with other parameters. 

Overall, we see that initial frequency has a large impact on whether an allele becomes fixed or extinct. Without a high initial frequency, it is very unlikely a new allele that appears in a population will last without extraordinary conditions. 

## Fixation

Next, let's determine the relationship between the number of generations it takes on average to get an allele to fixation, and the population size. 
```{r, fig.align='center'}
population <- c(10, 100, 1000, 10000)
generations <- 100
matrix <- matrix(0,nrow=generations,ncol<-length(population))

for(i in 1:generations){
  fixation <- c()
  for(j in population){
    pA <- c()
    pA[1] <- 0.5
    k <- 1
    while ((pA[k] < 1) & (pA[k] > 0)) {
      pA[k+1] <- rbinom(n=1,size=2*j,prob=pA[k])/(2*j)
      k <- k+1
    }
    fixation <- c(fixation,k)
  }
  matrix[i,] <- fixation
}
means <- colMeans(matrix)
plot(population, means, xlab = "Population Size", ylab = "Fixation Time")
abline(lm(means~population))
```
From the graph, it seems that there is a linear correlation between population size and the number of generations it takes a new allele to reach fixation. 